Untitled

### 2016-11-27 Response_to_reviewers.Rmd

We have added context to this choice and the implications.  Using long term records reported in Baker et al (1985) and Burnett et al. (2010) we indicate that the long term trend is overall warming in the region.  We had previously used PDSI since this was the only available long term record we had access to.  This change resulted in the inclusion of the folowing senteces in the Methods:

*Baker et al. (1985) show long term increases in mean annual temperature at Ft. Snelling and other spatially proximate records, while corrected estimates from Fort Leavenworth in Kansas Burnette et al. (2010) show increasing temperatures in all seasons, with faster increases during winter (0.10oC decade-1) than during summer (0.07oC decade-1).  This implies that historical estimates may be biased by +0.35oC for $t_{max}$, by +0.5oC for $t_{min}$ and +0.15oC for $t_{diff}$.*

and in the Results:

*When long term records (Burnette et al., 2010; Baker et al., 1985) are considered, $\Delta t_{min}$ increases to approximately `round(clim_change[4], 1) + 0.5`^o^C, $\Delta t_{max}$ to `round(clim_change[2], 1) + 0.35`^o^C and $\Delta t_{diff}$ to `round(clim_change[2], 1) - 0.15`^o^C.*

@@

We have added the sentence: *"The Modern climate data uses a climate normal that encompases the period of observation  (2000 to 2015) for the FIA data used in this study."*

![](figures/boxplot_corrected_all.png)

> A third issue is the role of climate variability over the 1895 to 2014 time period.  Decadal oscillations are common in this period in many parts of the US.  Trees are likely integrating over 2 or more of these climate cycles.  SDMs should be done to best represent the climate and variability that selected for tree presence when the trees were measured.  Your analysis would be strengthened with statistical analysis of climate patters from 1895 to present for your predictor variables and an analysis of which time period produces the best SDMs.

I believe that this is addressed in the response to the prior comment.  We have added a LOESS curve to Figure 1, we have selected a normal that reflects the period of sampling, and have provided additional support to the choice of the modern normal.

@@

See the following note.

> 365-370 These sentences are a bit of bait and switch.  Yes, readers will agree that SDM based on FIA may not fully represent the realized niche of tree species.  However, some of your argument hinges on the magnitude of the error that is made by using FIA.  I think your paper way over states that error.  The paper would have more impact if you were careful to emphasize the limitations of the data sets for this analysis and say that the degree to which FIA leads to SDM error is not knowable with available data sets.

I believe the reviewer possibly overstates the case for limitations.  There is strong support for changes since EuroAmerican settlement, both in vegetation-climate relationships, in the spatial extent of vegetation cover, and in pollen-vegetation relationships in the region, indicating a long-term shift in vegetation representation in climate space.  However, we have qualified the statements made in this section, and added a short paragraph pointing to the potential effects of the choice of climate normal as well.

*Some of these apparent shifts in climate niches may be due to differences in sampling design rather than to actual shifts in species distributions. In particular, the FIA dataset may be underrepresenting tree distributions in unforested or semi-forested regions. If so, estimates of forest loss (`fig_nums('gain_loss', display='cite')`) and niche shifts (Figures 4 & 5) may be overestimated, in particular along the warmest and driest region of the tree distributions in the Midwest. Regardless of cause, the differences between the PLS and FIA tree distributions are important for distributional modeling because FIA data provides the most detailed and most widely used source of tree distributional data over the entire contiguous United States [@wang2016landscape;@nieto2015close;@bell2014early]. Hence, any incompleteness in the representation of realized climate niches based on FIA data that are affecting this study should also affect ecological assessments that are based on FIA data. As improvements in remote sensing and statistical reconstructions of past forested landscapes improve it may become possible to fully explore this issue, but at present the extent of the problem is unclear, although this contribution points to possible effects of incomplete sampling.*

@@

> 3) Again, how the effects land use and exotic pathogens were included in the analyses?

> 4) The 2×2 factorial design with pre-settlement and modern vegetation and climates (VHCH, VHCM,VMCH,VMCM) was very confusing (Figure 2a). The historical distribution data for VHCH scenario were derived from PLS and the contemporary distribution data for VMCM were derived from US FIA data. Then, what were the data for VHCM and VMCH scenarios. It was impossible to get the data for VHCM and VMCH scenarios, right? If my understanding was wrong, then the more detailed information was needed to better describe the factorial design and data used for each scenario.

We have attempted to clarify this description by re-writing this paragraph:

*The total shift ($d_{tot}$) is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`fig_nums("fourPanel", display = "cite")`a: top left to bottom right). These conditions are observed or recorded in data.  The conditions $V_{M}C_{H}$ and $V_{H}C_{M}$ are synthetic, generated by superimposing historical climate patterns on modern vegetation distributions and by superimposing modern climate on historical distributions.  The effect of climate change on shifting species PDFs is described as $d_{c}$, the difference between PDFs using the early climate normals ($C_{H}$) and modern climate normals ($C_{M}$) overlaid on PLS vegetation (e.g., $V_{H}C_{H}$ versus $V_{H}C_{M}$; `fig_nums("fourPanel", display = "cite")`a; left column). We interpret PDF changes estimated from vegetation change ($V_{H}$ to {$V_{M}$) superimposed on  historical climate (*e.g.*, $V_{H}C_{H}$ versus $V_{M}C_{H}$; modeled by a shift along the top row of Figure 2a) to indicate change in vegetation attributed to land use change, or, more precicely, the land use change associated with EuroAmerican settlement.  We describe the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ as $d_{v}$.*

### 2016-12-02 Response_to_reviewers.Rmd

---
bibliography: goringetal_references.bib
csl: ecology.csl
output:
  word_document: 
    reference_docx: word_template.docx  
  pdf_document:
    pandoc_args: -V geometry:vmargin=1in -V geometry:hmargin=1in
  md_document:
    variant: markdown_github
pandoc_args: -H margins.sty
---

@@

# Response to Reviewers

Dear Dr. Thrall:

We submit for your consideration the revised manuscript "Effect of historical land-use and climate change on tree-climate relationships in the upper Midwestern United States."

We have carefully revised the manuscript in response to reviewer comments and have paid particular attention to the potential limitations associated with 1) comparing the PLS and FIA data and 2) choice of climate normal.  We completed several new analyses to test the sensitivity of our results to choice of climate normal and created a separate subsection within the Discussion assessing the robustness of our findings.

We have edited the manuscript for clarity and to respond to the other points raised by the reviewers.  We include below a point-by-point response to all reviews.  The revised version has XX  words and 6 figures and tables.  We have expanded the Supplementary Information to show several additional analyses.

We thank you and especially the reviewers for their thoughtful and constructive criticism – we believe that the manuscript has greatly improved as a result.  Thank you too for extending the deadline as requested.

Sincerely,

Simon Goring
Jack Williams

## Editor

*Editors Comments for the Author(s): The reviews highlight the importance and novelty of the main idea - that past land-use complicates our estimates of forest community responses to climate change. However, they express concern about the way that the results are presented. There are many suggestions for clarification that should be addressed in revision. In particular, it's important to address the limitations of comparing the two data types, as this was a concern expressed by each reviewer.*

Thank you.  Please see responses below.

@@

## Reviewer 1:

> The manuscript written by Goring and Williams (“Effect of historic land-use and climate change on tree-climate relationships in the northern United States”; ELE-00978-2016) addresses an important topic in global change research: what is the relative influence of climate vs. land-use change on contemporary species distributions. The question is relevant as the degree to which the two drivers are both important indicates potentially major errors in future projections previously published using species distribution models (SDMs). The authors present quite a clever method for addressing the issue: i.e., building SDMs with both historical and contemporary vegetation and climate data. This paper provides novel perspective concerning how our understanding of the role of climate change in vegetation change is still limited. The paper is exceedingly well written and the conclusions seem well-supported by the methods and results.

We thank the reviewer for the positive feedback and constructive criticisms.  We address specific points below.

@@

> My primary critique (which I admit is relatively minor) is that I found the logic revolving around the “confounding effects” and how the results relate to that a bit difficult. In particular, there likely needs to be a bit of clarification around the concept and a justification for why treating two non-significant effects and a set of mixed effects (one significant and one not so) as equivalent within this framework.

To address this point, we’ve expanded Table 1 to indicate the individual terms (dV and dL) and their significance.  This lets the reader see for themselves what goes into 'confounding'.  We also added to the Introduction (4th paragraph) the sentence “These confounding effects may make it more difficult to detect climatic forcing of historical species distributions.”  As it turns out, in the primary analyses presented here, all instances of confounding involve a significant change in climate and no significant change in vegetation (Table 1)

@@

> 347-350 Is it accurate to infer that ag has “largely eliminated open forests” which does not sample low density forests? It seems likely that most of these tree species are present in at least low densities in the areas where your maps show they are not present.

As above, we believe that these patterns are real.  We extend this point however beginning at **l**371, where we discuss the potential use of isolated trees or trees unsampled in the FIA data.

> Figure 2 b-e, The axes should be labeled and units indicated.  Consider reformatting these figures. The current versions make visual separation of the lines hard to detect.

We have updated this figure to make the comparison clearer.

![](figures/figure_two.png)

@@

> Firstly, it is not possible to get the distribution data under VMCH. Secondly, the distribution changes over the last two centuries were caused by many factors including land use change, disturbances (e.g., pathogen outbreaks, fire, harvest), and endogenous succession processes. Why the difference VH and VM could represent the land use change effects? This was too arbitrary. A stronger rationale is needed.

We believe that our argument is valid, but may have been unclear in the framing.  First, we believe that attributing the distributional changes from VH to VM to land use is justified.  Shulte et al (2007) and Rhemtulla et al (2009) have used similar language to explain the set of changes associated with EuroAmerican settlement. Initial clearance for agricultural and urban development was contemporaneous with massive logging operations, and, in the late 1800s, widespread fire originiating from human sources.  Given this, and the fact that very little "old growth" forest remains, modern forests are indellibly imprinted with the signal of human land use, whether agricultural conversion or timbering.  The IPCC considers forestry to be a component of land use.  The Peshtigo firestorm, the Michigan fires of 1871 and 1881 were driven by slash buildup following logging.  While sucession processes occurred since clearance and settlement, these were directly affected by the initial land clearance.  For example, Rhemtulla et al. (2009) indicate that the loss of hemlock was caused both by logging, and recruitment failure as the result of the loss of seed source because of the extensive logging, thus both cause & effect are rooted in EuroAmerican settlement.

> The authors should use present tense not past tense in the methods and results sections.

We have changed this throughout.  We except descriptions of methods applied in earlier papers, or in the preparation of data prior to the current study (for example, when summarizing the methods of Goring et al. 2016).
